Testing and Isolation Efficacy: Insights from a Simple Epidemic Model

### 2021-06-09 SIR_manuscript.tex
meeting with Ben comments

The basic reproduction number, $\Rnum$, was calculated by using the next-generation matrix method \citep{van2002reproduction}. We present $\Rnum$ as
\begin{equation}
\label{R0}
\Rnum= \frac{\beta}{\gamma} \left(1-\Delta\right), 
\end{equation}
where the term $\beta/\gamma$ is the classical basic reproduction number for a SIR model without testing and isolation \citep{keeling2011modeling}, and $\Delta>0$ is the effectiveness of control due to testing and isolation processes defined as 
\begin{equation}
  \label{eq:del4a}
  \Delta= \frac{1}{C N_0}\big(C_1 S\_u^*+(C_2(1-\theta\_w)+C \theta\_w)S\_n^*\big),
\end{equation}
where
\begin{align}
\label{eq:C12}
C_{\phantom 1} &= (\omega+\gamma) \Big(\gamma(\omega+\gamma)+(\gamma+\omega p_I)\hat F_I\Big),\\
C_1 &= (\omega+\gamma)(\theta\_w \gamma+\theta\_c \omega p_I) \hat F_I,\\
C_2 &= \Big( \omega\gamma(1+p_I)\hat F_I+\gamma^2(\omega+\gamma+\hat F_I)\Big)\theta\_w + \omega^2p_I\hat F_I\theta\_c.
\end{align}
(Appendix~\ref{app:R0} gives a detailed derivation of these expressions).
The explicit formula of $\Rnum$ enables us to study the effects of testing and isolation parameters on $\Rnum$ both analytically and via numerical solutions.
We are specifically interested in parameters that could be manipulated by public health policy: isolation efficacy, $\theta\_c$ and $\theta\_w$; \percap testing intensity, $\rho$; and the rate of test return, $\omega$. In particular, we look at the partial derivatives of $\Delta$ with respect to these parameters (App.~\ref{app:rho}). 
We derived general expressions for these derivatives. However, we analyzed the effect of $\omega$ on $\Rnum$ for the special case of low testing intensity. Specifically, by making the restriction $\rho \ll 1$, we are able to Taylor-expand $\Delta$ at $\rho=0$ and analyze the resulting simplified derivatives to illustrate a surprising non-monotonic relationship between $\Rnum$ and $\omega$.

The analytical calculation of the next-generation matrix and simplification of the $\Rnum$ expression was carried out in Maple \citep{maple14}. We used R \citep{r} for numerical solutions, in particular for plotting the contours of $\Delta$ (Eq.~\eqref{eq:del4a}) over a range of parameters.
We computed the values and contours of Delta at both low (\fref{pan}) and high (\fref{pan2}) testing intensities, and for both random testing ($w_S=w_I=w_R=1$) and targeted testing ($w_S=0.3$; $w_I=w_R=1$).

@@

In this paper, we have defined and analyzed a simple compartmental model that combines epidemiological dynamics (as defined by a simple SIR model) with the dynamics of testing and isolation. Our model is a caricature --- it models the most basic feedbacks between epidemic and testing processes, but does not attempt to incorporate the many known complications of COVID-19 epidemiology (e.g. exposed, presymptomatic, and asymptomatic compartments [REFS]\ali{\citep{kain2021chopping} OK?}; time-varying testing rates [REFS?]\ali{\citep{jarvis2021temporal} OK?}; behavioural dynamics [REF Weitz/Park/Dushoff] \ali{\citep{weitz2020awareness} OK?}). Thus, it is most appropriate for assessing the \emph{qualitative} phenomena that arise from the interactions between epidemiology and testing, rather than for making quantitative predictions or guiding pandemic responses.

@@

\bmb{Leftover comments:
\begin{itemize}
\item should emphasize (at some appropriate point in text) that $\Delta$ is independent of $\beta$ (and hence of $\Rnum$ if we are thinking primarily of holding $\gamma$ constant and varying $\beta$) for the small-$\rho$ (Taylor-expansion) case, see eq. ...
\item it would be good to have a few more testing refs: Bergstrom et al., there might be some other stuff we can recycle from the lab meeting review; Friston's stuff deserves to be mentioned because it does have explicit testing flows too. We should try not to ignore/piss anyone off who has already done stuff
\end{itemize}
}

@@

% %%%%%%%
\subsection{rate of returning tests} \label{app:omega}
{\bf The third goal} is to explore how changes in the rate of test return $\omega$ affects $\Rnum$. Mathematically we would like to verify the sign of $\pder\Rnum{\omega}$, which specifically depends on $\pder\Delta{\omega}$. We use
the linearization of $\Rnum$ around $\rho=0$ to show that there a non-monotonic relationship between $\Rnum$  and $\omega$. The Taylor expansion of $\Delta$ at $\rho=0$ is
\begin{equation}
\label{eq:lin}
\Delta = \frac{\rho}{\omega \gamma(\omega+\gamma)} \Big(
\frac{w_I}{w_S} \omega^2 p_I \theta\_c+(\frac{w_I}{w_S}+1) \gamma\omega\theta\_w +\gamma^2 \theta\_w(2-\theta\_w)
\Big). 
\end{equation}
This results in
\begin{equation}
\label{eq:dlindo}
\pder\Delta{\omega} = \frac{\rho}{\omega^2 (\omega+\gamma)^2} \Big(
(p_I \frac{w_I}{w_S}\theta\_c -(1+\frac{w_I}{w_S})\theta\_w)\omega^2+2\theta_w \gamma(\theta_w-2) \omega+\theta_w \gamma^2 (\theta_w-2)
\Big),
\end{equation}
around $\rho=0$.
% \ali{I stoped here!}

>>>>

The basic reproduction number, $\Rnum$, was calculated by using the next-generation matrix method \citep{van2002reproduction}. We present $\Rnum$ as
\begin{equation}
\label{R0}
\Rnum= \frac{\beta}{\gamma} \left(1-\Delta\right), 
\end{equation}
where the term $\beta/\gamma$ is the classical basic reproduction number for a SIR model without testing and isolation \citep{keeling2011modeling}, and $\Delta>0$ is the effectiveness of control due to testing and isolation processes defined as 
\begin{equation}
  \label{eq:del4a}
  \Delta= \frac{1}{C N_0}\big(C_1 S\_u^*+(C_2(1-\theta\_w)+C \theta\_w)S\_n^*\big),
\end{equation}
where
\begin{align}
\label{eq:C12}
C_{\phantom 1} &= (\omega+\gamma) \Big(\gamma(\omega+\gamma)+(\gamma+\omega p_I)\hat F_I\Big),\\
C_1 &= (\omega+\gamma)(\theta\_w \gamma+\theta\_c \omega p_I) \hat F_I,\\
C_2 &= \Big( \omega\gamma(1+p_I)\hat F_I+\gamma^2(\omega+\gamma+\hat F_I)\Big)\theta\_w + \omega^2p_I\hat F_I\theta\_c.
\end{align}
(Appendix~\ref{app:R0} gives a detailed derivation of these expressions).
The explicit formula of $\Rnum$ enables us to study the effects of testing and isolation parameters on $\Rnum$ both analytically and via numerical solutions.
We are specifically interested in parameters that could be manipulated by public health policy: isolation efficacy, $\theta\_c$ and $\theta\_w$; \percap testing intensity, $\rho$; and the rate of test return, $\omega$. In particular, we look at the partial derivatives of $\Delta$ with respect to these parameters (App.~\ref{app:rho}). 
We derived general expressions for these derivatives. However, we analyzed the effect of $\omega$ on $\Rnum$ for the special case of low testing intensity. Specifically, by making the restriction $\rho \ll 1$, we are able to Taylor-expand $\Delta$ at $\rho=0$, use the linear approximation with respect to $\rho$ and analyze the resulting simplified derivatives to illustrate a surprising non-monotonic relationship between $\Rnum$ and $\omega$.

The computations in this paper, specifically, calculation of the next-generation matrix and simplification of the $\Rnum$ expression, were performed in Maple\textsuperscript{\texttrademark} \citep{maple14}. We used R \citep{r} for numerical solutions, in particular for plotting the contours of $\Delta$ (Eq.~\eqref{eq:del4a}) over a range of parameters.
We computed the values and contours of Delta at both low (\fref{pan}) and high (\fref{pan2}) testing intensities, and for both random testing ($w_S=w_I=w_R=1$) and targeted testing ($w_S=0.3$; $w_I=w_R=1$).

@@

In this paper, we have defined and analyzed a simple compartmental model that combines epidemiological dynamics (as defined by a simple SIR model) with the dynamics of testing and isolation. Our model is a caricature --- it models the most basic feedbacks between epidemic and testing processes, but does not attempt to incorporate the many known complications of COVID-19 epidemiology (e.g. exposed, presymptomatic, and asymptomatic compartments [REFS]\ali{\citep{kain2021chopping} OK?}; time-varying testing rates [REFS?]; behavioural dynamics [REF Weitz/Park/Dushoff] \ali{\citep{weitz2020awareness} OK?}). Thus, it is most appropriate for assessing the \emph{qualitative} phenomena that arise from the interactions between epidemiology and testing, rather than for making quantitative predictions or guiding pandemic responses.

@@

\bmb{Leftover comments:
\begin{itemize}
\item should emphasize (at some appropriate point in text) that $\Delta$ is independent of $\beta$ (and hence of $\Rnum$ if we are thinking primarily of holding $\gamma$ constant and varying $\beta$) for the small-$\rho$ (Taylor-expansion) case, see Eq.~\ref{eq:lin}.
\item it would be good to have a few more testing refs: Bergstrom et al., there might be some other stuff we can recycle from the lab meeting review; Friston's stuff deserves to be mentioned because it does have explicit testing flows too. We should try not to ignore/piss anyone off who has already done stuff
\end{itemize}
\ali{:}
Testing dynamics has been previously modeled in the context of epidemiology.
\cite{bergstrom2020frequency} modeled repeated random testing of isolated populations and provided some analytical results.
\citep{friston2021testing} model the effects of self-isolation on testing and tracking.  
 Comparing to these works, our modeling execise is novel. 
}

@@

% %%%%%%%
\subsection{rate of returning tests} \label{app:omega}
{\bf The third goal} is to explore how changes in the rate of test return $\omega$ affects $\Rnum$. Mathematically we would like to verify the sign of $\pder\Rnum{\omega}$, which specifically depends on $\pder\Delta{\omega}$. We use
the linearization of $\Delta$ around $\rho=0$ to show that there a non-monotonic relationship between $\Rnum$  and $\omega$. The linear term in the Taylor expansion of $\Delta$ at $\rho=0$ is
\begin{equation}
\label{eq:lin}
\Delta = \frac{\rho}{\omega \gamma(\omega+\gamma)} \Big(
\frac{w_I}{w_S} \omega^2 p_I \theta\_c+(\frac{w_I}{w_S}+1) \gamma\omega\theta\_w +\gamma^2 \theta\_w(2-\theta\_w)
\Big). 
\end{equation}
This results in
\begin{equation}
\label{eq:dlindo}
\pder\Delta{\omega} = \frac{\rho}{\omega^2 (\omega+\gamma)^2} \Big(
(p_I \frac{w_I}{w_S}\theta\_c -(1+\frac{w_I}{w_S})\theta\_w)\omega^2+2\theta_w \gamma(\theta_w-2) \omega+\theta_w \gamma^2 (\theta_w-2)
\Big),
\end{equation}
around $\rho=0$.
% \ali{I stoped here!}
